A supertree pipeline for summarizing phylogenetic and taxonomic information for millions of species

### 2017-01-10 doc/paper1-submission/paper5.tex
Alter terminology paragraph vis a viz Luay Nakleh.

Let $\summaryTree$ denote a supertree, and $T_{i}$ denote the $i$th
input tree. The set of taxa that are mapped to the tips of the tree
$T_{i}$ is $\mathcal{L}(i)$. $\summaryTree(i)$ denotes the summary
tree induced by tip nodes that are mapped to taxa in $\mathcal{L}(i)$
and the most recent common ancestor of those leaves, and any other
node that is an ancestor of some but not all of these leaves. We say
that edge $j$ of the supertree is compatible with an input tree,
$T_{i}$ if edge $j$ either is not included in the induced tree $\summaryTree(i)$
or none of the edges in $T_{i}$ are in conflict with edge $j$ in
the induced tree.

We can consider whether or not a node in an input tree is displayed
by $\summaryTree$. For any such node $j$ there is a set of taxa
that are mapped to the tips that descend from the node. This set of
taxa is the ``cluster'' of taxa corresponding to node $j$; it can
be denoted $\mathcal{L}(i,j)$. It will also be referred to as the
``include set'' of the node. The ``exclude set'' of $j$ in $T_{i}$
is the set of taxa in $\mathcal{L}(i)$ but not in $\mathcal{L}(i,j)$.
If the cluster of taxa for any supertree node in $\summaryTree(i)$
is identical to $\mathcal{L}(i,j)$, then we say that $\summaryTree$
displays node $j$ of $T_{i}$. We say that the summary tree \emph{displays}
edge $j$ if the summary tree displays the child node of edge $j$.
Operationally, we can find the most recent common ancestor (MRCA)
node of $\mathcal{L}(i,j)$ in $\summaryTree$; the summary tree displays
$j$ if and only if that MRCA node is not an ancestor of any member
of the exclude set of $j$. We say that node $k$ of the summary tree
is \emph{supported by} node $j$ in $T_{i}$ if the summary tree displays
node $j$, but if we contracted edge that separates $k$ from its parent then the modified summary
tree would no longer display node $j$.

@@

Propinquity was designed to function as a part of the Open Tree of
Life software architecture, so the first few steps of the pipeline
involve transforming artifacts from that project into a set of rooted
trees and a phylogenetic taxonomy. The phylesystem API \citep{McTavishEtAt2015}
of Open Tree allows users to curate published estimates of trees and
create ranked collections of these trees. Early steps in the propinquity
pipeline manipulate the phylogenetic input trees to improve their
usability and reliability. The first steps of the pipeline (see Figure
\ref{fig:pipeline}) collect a list of trees to include (in the \texttt{phylo\_input}
subdirectory) and store copies of these files (in the \texttt{phylo\_snapshot}
subdirectory) to make it easier to replicate the operation (because
the collection of trees and the tree files change due to curation).
\begin{figure}
\begin{centering}
\includegraphics[height=0.75\paperheight]{pipeline-tools.pdf}
\par\end{centering}
\caption{Organization of the \emph{propinquity} pipeline. Each colored pentagon
labels a program (blue for otcetera-based tools and red for python
scripts in the propinquity or peyotl repository) that performs the
important operations in each step; the number before the tool name
refers to the section in this paper that describes the operation.
The output of each step corresponds to a subdirectory of the propinquity
system which will hold the output artifacts for the step. Ovals are
resources that are required (OTT and Open Tree's phylesystem repository).
White pentagons are user-controlled inputs.}
\label{fig:pipeline}
\end{figure}

>>>>

Cutting any edge \textbf{$j$ }of a rooted tree induces a bi-partition
$S(j)$ of the leaf taxa into two connected groups $S_{1}(j)$ and
$S_{2}(j)$. Such a bi-partition is called a rooted split and written
$S(j)=S_{1}(j)|\bullet S_{2}(j)$. Here $S_{1}(j)$ is called the ``cluster'' of $j$, and contains tip taxa on the side of edge $j$ that does not contain the
root. $S_{2}(j)$ contains the tip taxa on the side of edge $j$ that
contains the root. We refer to $S(j)$ as a \emph{rooted} split because
the right side of the split implicitly contains the root node, as indicated
by the $\bullet$.

For any two rooted splits $A=A_{1}|\bullet A_{2}$ and $B=B_{1}|\bullet B_{2}$,
we say that $A$ \emph{displays} $B$ if $B_{1}\subseteq A_{1}$ and
$B_{2}\subseteq A_{2}$. We also say that $A$ and $B$ \emph{conflict}
if none of $A_{1}\cap B_{1}$, $A_{1}\cap B_{2}$, and $B_{2}\cap A_{1}$
are empty\footnote{The intersection of the right sides of the splits is considered to
be non-empty even if $A_{2}\cap B_{2}$ is empty, because the right
sides implicitly contain the root.}. If $A$ and $B$ do not conflict, then we say that they are \emph{compatible}\footnote{It is possible to construct a rooted tree that displays both $A$
and $B$ if and only if $A$ and $B$ are compatible}. If $A$ displays $B$, then $A$ and $B$ must be compatible, since
$A_{1}\cap B_{2}$ and $B_{1}\cap A_{2}$ must be empty. We adopt
the shorthand of referring to an edge $k$ when we mean the split
$S(k)$. So, when we say that edge $j$ conflicts with edge $k$,
we mean that $S(j)$ conflicts with $S(k)$.

Let $\summaryTree$ denote a supertree, and $T_{i}$ denote the $i$th
input tree. We say that $\summaryTree$ displays an edge $j$ of $T_{i}$
if any edge of $\summaryTree$ displays $j$. We say that $\summaryTree$
conflicts with a edge $j$ of $T_{i}$ if any edge of $\summaryTree$
conflicts with $j$; otherwise $\summaryTree$ is compatible with
$j$. We say that edge $k$ of $\summaryTree$ is \emph{supported
by} edge $j$ of $T_{i}$ if $\summaryTree$ displays $j$, but $\summaryTree$
would not display $j$ if we contracted edge $k$. We also adopt the
shorthand of referring to edges by their tipward endpoint. By combining
this with the above shorthand, we could say that a node $j$ conflicts
or does not conflict with another node $k$.

The set of taxa that are mapped to the tips of the tree $T_{i}$ is
$\mathcal{L}(i)$. $\summaryTree(i)$ denotes the summary tree induced
by tip nodes that are mapped to taxa in $\mathcal{L}(i)$ and the
most recent common ancestor of those leaves, and any other node that
is an ancestor of some but not all of these leaves.

@@

Propinquity was designed to function as a part of the Open Tree of
Life software architecture, so the first few steps of the pipeline
involve transforming artifacts from that project into a set of rooted
trees and a phylogenetic taxonomy. The phylesystem API \citep{McTavishEtAt2015}
of Open Tree allows users to curate published estimates of trees and
create ranked collections of these trees. Early steps in the propinquity
pipeline manipulate the phylogenetic input trees to improve their
usability and reliability. The first steps of the pipeline (see Figure
\ref{fig:pipeline}) collect a list of trees to include (in the \texttt{phylo\_input}
subdirectory) and store copies of these files (in the \texttt{phylo\_snapshot}
subdirectory) to make it easier to replicate the operation (because
the collection of trees and the tree files change due to curation).
\begin{figure}
\begin{centering}
\includegraphics[height=0.70\paperheight]{pipeline-tools.pdf}
\par\end{centering}
\caption{Organization of the \emph{propinquity} pipeline. Each colored pentagon
labels a program (blue for otcetera-based tools and red for python
scripts in the propinquity or peyotl repository) that performs the
important operations in each step; the number before the tool name
refers to the section in this paper that describes the operation.
The output of each step corresponds to a subdirectory of the propinquity
system which will hold the output artifacts for the step. Ovals are
resources that are required (OTT and Open Tree's phylesystem repository).
White pentagons are user-controlled inputs.}
\label{fig:pipeline}
\end{figure}
